---
title: "Computer Practical 1"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1 and Q2


```{r cars}
n <- 13
p <-0.31
# we will carry out 100 trials in which each trial we observe 10 Binomial(n,p)
sampleSize <- 10
trials <- 100
# varHat stores the true variance of the MLE of p
varHat <- (1/n) * p * (1-p)
print(paste("Variance of pHat(Y) is " ,varHat))
# phat is the MLE of p function
phat <-function(Y) {
  Y/n
}
# we generate 1000 binomials and organise them into 10 rows of 100 
# this means each column is a trial
xValues <- rbinom(sampleSize*trials, n, p)
xSamples <- matrix(xValues, nrow=sampleSize) # 10x100
# we pass each row into phat to get the MLE function transformation 
pHatSamples<-apply(xSamples, 1, phat) # it becomes 100x10
# we then get the sample variance of each row 
pHatVar <- apply(pHatSamples, 1, var) # stays 100x1
hist(pHatVar, xlab = "Variance")
abline(v = varHat, col='red')

```

## Q3 (Q4 and Q5 done in working out section)


```{r pressure}
year.data <- read.csv("year_data.csv")
knitr::kable(year.data)
n1 <- sum(year.data[year.data$clinic==1,]$births)
y1 <- sum(year.data[year.data$clinic==1,]$deaths)
n2 <- sum(year.data[year.data$clinic==2,]$births)
y2 <- sum(year.data[year.data$clinic==2,]$deaths)
print(paste("The number of births in clinic 1 is ", n1))
print(paste("The number of deaths in clinic 1 is ", y1))
print(paste("The number of births in clinic 2 is", n2))
print(paste("The number of deaths in clinic 2 is", y2))
print(paste("The ML estimates of p1hat(y1) is y1/n1 which is 1989/20042 which is",y1/n1))
print(paste("The ML estimates of p2hat(y2) is y2/n2 which is 691/17791 which is " , y2/n2))
```

# Q6 First Part (second part done in working out section)

```{r second}
month.data <- read.csv("month_data.csv")
# removes NA data items 
month.data <- month.data[!is.na(month.data$births),]
month.data$rate <- month.data$deaths/month.data$births
month.data$date <- as.Date(month.data$date)
intervention.date <- as.Date("1847-05-15")
plot(month.data$date, month.data$rate, pch=20, main="Mortality rate by month", 
     xlab="Date; red line indicated start of intervention period", ylab="Rate ")
abline(v=intervention.date, col="red")

before.interventions <- month.data[month.data$date < intervention.date,]
after.intervention <- month.data[month.data$date > intervention.date,]

n1 <- sum(before.interventions$births)
y1 <- sum(before.interventions$deaths)
n2 <- sum(after.intervention$births)
y2 <- sum(after.intervention$deaths)
print(paste("The number of births before intervention is ", n1))
print(paste("The number of deaths before intervention is ", y1))
print(paste("The number of births after intervention is", n2))
print(paste("The number of deaths after intervention is", y2))

print(paste("The ML estimates for the mortality rates before intervention is y1/n1 which is " , y1/n1))
print(paste("The mortality rates after intervention is y2/n2 which is ", y2/n2))
